---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# `forplot` 

A package to generate forest plots.

## Installation

`forplot` can be installed from github:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("CTU-Bern/forplot")
```

Note that `remotes` treats any warnings (e.g. that a certain package was built under a different version of R) as errors. If you see such an error, run the following line and try again:

```{r remotes-error, eval = FALSE}
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true")
```

## Overview
`forpplot` and it's main function fplot needs a data frame with specific column names. 

- vlabel: a chr column with the variable labels printed on the left of the forest plot 
- nx: any number of chr or num columns numbered sequentially (i.e. n1, n2, n3, ...). Printed in that seqence after the 
	column label. May contain the number of observations and/or summary per group.
- beta, beta_lci, beta_uci: three num columns with point estimates and confidence interval to be plotted as forest.
- beta_format: optional chr column with formatted text to be prin§ted along forest, generated from beta if not given.
- px: any number of chr or num columns numbered sequentially (i.e. p1, p2, p3, ...), printed on the very right of the plot.
	May contain p-values.

The package includes an example dataset with 10 variables:

```{r}
# load package
library(forplot)

# demonstration data
data(forplotdata)

forplotdata
```

The nx columns included the number of observations and descritpives (mean (sd)) for each group, beta is a mean difference, p1 the p-value.

The minimal plot only includes a label and the forest and needs columns vlabel, beta, beta_lci, beta_uci.

```{r, fig.height=3.5, fig.width=5}
fplot(dat=forplotdata[,c("vlabel","beta","beta_lci","beta_uci")])
```

We can add the nx and px columns:
```{r, fig.height=3.5, fig.width=8}
fplot(dat=forplotdata)
```

Set more sensible widths and heights.
The widths has to be the same length as the number of columns for the input 
	data plus two for the left and right margin.
The length is usually three, header, body and footer.	

```{r, fig.height=3.5, fig.width=8}
lwidths<-c(0.05,0.5,0.2,0.8,0.2,0.8,1.2,1.2,0.5,0.05)

lheights<-c(0.14,1,0.08)

fplot(dat=forplotdata,lwidths=lwidths,lheights=lheights)
```

A header can be given using a character vector the same 
	length as the number of columns of the input data, which then 
	places it above the columns

```{r, fig.height=3.5, fig.width=8}
header<-c("","Group1\nN","Group0\nmean (sd)","Group2\nN","Group2\nmean (sd)",
	"Mean difference\n95% CI","","P-value")

fplot(dat=forplotdata,lwidths=lwidths,lheights=lheights,header=header)
```

The header can also be placed at any x/y position using a list 
	with one element per header line.

```{r, fig.height=3.5, fig.width=8}
header<-list(list(y=0.7,
	text=c("Group1","Group2","Mean difference (95% CI)","P-value"),
	x=c(0.10,0.32,0.7,0.98)),
	list(y=0.3,text=c("N","mean (sd)","N","mean (sd)"),
	x=c(0.07,0.18,0.28,0.38)))

fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights)
```

A label can be placed below the x-axis using option xtitle.
Option ref adds a vertical reference line and xlim specifies the 
	limits for the axis.

```{r, fig.height=3.5, fig.width=8}
xtitle<-list(x=0.86,y=0.2,textl="Group 1 better  ",textr="  Group 2 better")

fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights,
	xtitle=xtitle,ref=list(x=0),xlim=c(-1,0.5))
```

Note that arrows are shown if the limits of the CIs are not included within xlim.

There are further options for the reference line and we can shift the x-axis
  if the gap at the bottom is too large.
	
```{r, fig.height=3.5, fig.width=8}
xtitle<-list(x=0.86,y=0.6,textl="Group 1 better  ",textr="  Group 2 better")

fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights,
	ref=list(x=0,col=2,extend=2),
	xtitle=xtitle,xlim=c(-1,0.5),shift_xaxis=0.3,xlab_line=-0.8)

```

The points of the forest plots can also be formatted.

```{r, fig.height=3.5, fig.width=8}
ps<-list(pch=16,cex=rnorm(10,2,0.2),col=1)

fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights,
	ref=list(x=0,col=2,extend=2),
	xtitle=xtitle,xlim=c(-1,0.5),shift_xaxis=0.3,xlab_line=-0.8,
	ps=ps)

```

Lines at the top and bottom can be added.

```{r, fig.height=3.5, fig.width=8}
fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights,
  ref=list(x=0,col=2,extend=2),
	xtitle=xtitle,xlim=c(-1,0.5),shift_xaxis=0.3,xlab_line=-0.8,
	headline=2,bottomline=1)
```

If effect measures are on the log-scale (e.g. for odds ratios), option lscale can be used to indicate that the text should contain the exponentiated values. Axis ticks and labels have to be adapted by hand.

```{r, fig.height=3.5, fig.width=8}

xlab_text<-c(0.3,0.5,0.8,1.0,1.5)
xlab<-log(xlab_text)
xlim<-log(c(min(xlab_text),max(xlab_text)))

fplot(dat=forplotdata,header=header,lwidths=lwidths,lheights=lheights,
  ref=list(x=0,col=2,extend=2),
  lscale=TRUE,
	xtitle=xtitle,shift_xaxis=0.3,xlab_line=-0.8,
  xlim=xlim,xlab=xlab, xlab_text=xlab_text,
	headline=2,bottomline=1)

```


