---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# `forplot` 

A package to generate forest plots.

## Installation

`forplot` can be installed from github:

```{r gh-installation, eval = FALSE}

#install package using remotes
remotes::install_github("dcr-unibe-ch/forplot")

# load package
library(forplot)

```

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
devtools::load_all()

library(forplot)

```


## Usage
`forplot` contains two main functions to build a forest plot.

`genfobj` requires a *layout*, specifying the elements to be plotted and a data frame (*dat*) with the data to be plotted and 
produces a list of class **fobj**.

The **fobj** can be modified using a set of helper functions or directly by changing the list.

`plotfobj` produces the plot from the **fobj**.

## Example data 

The package includes three data sets:

- *forplotdata*: summary data with a 10 continuous variables
- *forplotdata_prop*: summary data with 10 binary variables
- *forplotdata_bp*: data with actual observation of *forplotdata* in a long format

```{r}
data(forplotdata)
forplotdata
```


```{r}
data(forplotdata_prop)
forplotdata_prop
```
```{r}
data(forplotdata_bp)
head(forplotdata_bp)
```

*forplotdata* has a text column with the variable names (*vlabel*), text columns *n[1-4]* with the number of observations and descriptives for each arm (either mean (sd) or n(%)), the formatted treatment effect (*beta_format*) and the p-value (*p1*), and
	three numeric columns to draw the forest (*beta*, *beta_lci* and *beta_uci*).

In *forplotdata_prop* there are two further numeric columns with the proportion in each arm (*prop1*, *prop2*) to draw a srip plot.	

*forplotdata_bp* has three columns with the actual observations of *forplotdata* with

- *value*: the variable value (numerical),
- *variable*: the variable name (factor), and 
- *arm*: the treatment arm (factor).


## Generating the **fobj** with `genfobj`

`genfobj` requires the input of a *layout* and a data frame *dat*.

*layout* must be a character vector with elements *t* (text), *f* (forest), *s[1-9]* (strip), or *b* (boxplot).

For each *t* element, *dat* must contain a single column, 
for each *f* element three columns (point estimate, lower confidence limit, upper confidence limit, in that order), and
for each *s[1-9]* the number of columns indicated in []. 

The order of the columns must correspond to the layout.

A *b* element does not need a column in *dat* but the specifcation of *obs*, a 
	data frame in a long format with columns
	*value* (the outcome value), *variable* (the outcome variable, safest as a factor to preserve the order in the plot), 
	and *arm* (the treatment arm, safest as a factor to preserve the order in the plot)


We want to use *forplotdata* and generate a plot with a text columns with the label, the descriptives, the formatted effect and the p-value, and a forest for the beta.

```{r}
fobj<-genfobj(dat = forplotdata, layout = c("t","t","t","t","t","t","f","t"))
```

The produced **fobj** is a list of length 5 with class *fobj*.
It includes those elements:

- *dat* and *obs*: the data 
- *setup*: the options from `genfobj` which has been given or assumed by default (i.e. *layout*, *lheights*, *lwidths*, *y.at*, *ylim*)
- *items*: a list of length from *layout* with the options for each item
- *header*: a list with the options for the header


```{r}
class(fobj)
names(fobj)
fobj$setup
length(fobj$items)
```
By default, text items include these options:

```{r}
names(fobj$item[[1]])

fobj$item[[1]]

```

*type* is the item type, *vname* is the variable in *dat* the item corresponds to, *plot* and *text* are the options used for plotting with base r.

The *f* items are more complex, including options for *axis*, *points* and *arrows* of the fores plots.

```{r}
names(fobj$items[[which(fobj$setup$layout=="f")]])

fobj$items[[which(fobj$setup$layout=="f")]]$type

fobj$items[[which(fobj$setup$layout=="f")]]$vname
```

For the variable name, the first of the three variables used to specify the forest is used.
Options for axis, point and arrows can be changed using helper functions (or directly in the list).
Examples or shown below.

## Plot the **fobj**

The **fobj** can be plotted using `plotfobj`

```{r, fig.height=3.5, fig.width=8}
plotfobj(fobj)
```

And with a bit nicer widths, which can be given as options in `genfobj` or just be changing the **fobj**:

```{r, fig.height=3.5, fig.width=8}
fobj$setup$lwidths <- c(0.3,0.4,0.6,0.4,0.6,1,1,0.5)

plotfobj(fobj)
```

Horizontal gridlines and stripes can be added with helper functions *gridlines* and *stripes*, which generate further elements in the fobj:

```{r}
fobj<-gridlines(fobj)

fobj<-stripes(fobj)

names(fobj)
```

```{r, fig.height=3.5, fig.width=8}
plotfobj(fobj)
```

## Modify the items

All items, grid lines and stripes can be modified using helper functions or within the list directly.

For *t* items helper function *t_options* allows to use all options used in R [graphics::text()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/text.html) function.
That can be done for a specific t item by using the number or column name of the item, or for all (by keeping item=NULL):


```{r, fig.height=3.5, fig.width=8}
fobj<-t_options(fobj = fobj, item = c("vlabel"), cex = 1.1, font = 2, col = "red", x=0.2, adj=0)

plotfobj(fobj)
```

For *f* items there are several helper function to modify the options for the different elements:
*f_axis*, *f_points*, *f_arrows*. Also here, all options from [graphics::axis()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/axis.html), [graphics::points()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/points.html) and [graphics::arrows()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/arrows.html) can be used.
As we do only have one forest item, we do not have specify the *item*.

```{r, fig.height=3.5, fig.width=8}
fobj<-f_axis(fobj = fobj, xlim = c(-1.1,0.2), at = c(-1,-0.6,-0.2,0.2), 
  tck = -0.03, mgp = c(2,0.5,0))

fobj<-f_points(fobj = fobj, pch = 16, cex = 1.5)

plotfobj(fobj)
```

With *f_refline*, a reference line can be added and with *f_direction* a label for the direction below the axis.
Note that the footer height has to be increased to fit the direction label.

```{r, fig.height=3.5, fig.width=8}
fobj<-f_refline(fobj, x = c(0, 0))

fobj<-f_direction(fobj, text = "A better    B better", line = 1.6)

fobj$setup$lheights[3]<-0.15

plotfobj(fobj)
```

To remove an added item, it can be set to NULL (or a new **fobj** could be generated):

```{r, fig.height=3.5, fig.width=8}
fobj$items[[which(fobj$setup$layout=="f")]]$refline<-NULL

plotfobj(fobj)
```

### Piping 

The modifying elements can be piped. For example to reproduce the **fobj** from above:
```{r, fig.height=3.5, fig.width=8}

fobj<-genfobj(dat = forplotdata,
  layout = c("t","t","t","t","t","t","f","t"),
  lwidth = c(0.3,0.4,0.6,0.4,0.6,1,1,0.5),
  lheight = c(0.1,1,0.15))

fobj |>
  gridlines() |>
  stripes() |>
  t_options(item = c("vlabel"), cex = 1.1, font = 2, col = "red", x=0.2, adj=0) |>
  f_axis(xlim=c(-1.1,0.2), at = c(-1,-0.6,-0.2,0.2), tck = -0.03, mgp = c(2,0.5,0)) |>
  f_points(pch = 16, cex = 1.5) |>
  f_refline(x = c(0, 0)) |>
  f_direction(text = "A better    B better", line = 1.6) |>
  plotfobj()

```

### Arrowheads

When using *xlim* via *f_axis*, x-axis limits that are narrower then the confidence limits can be set (which is usually not recommended as it can be misleading). 
The cap at the end of the confidence interval is then not plotted (as it is outside of the plotting region).
However, to make that clearer, directional arrowheads can be added using options *f_arrows*.
To have different heads on both sides, the arrow has to be drawn twice, with code 1 and 2. 
The helper function *f_cutarrows* helps with that:

```{r, fig.height=3.5, fig.width=8}
# Modify data to generate some very wide limits:
dat<-forplotdata
dat$beta_lci[1]<-c(-3)
dat$beta_uci[2]<-c(3)
dat[3,c("beta_lci","beta_uci")]<-c(-3,3)
dat[4,c("beta","beta_lci")]<-c(-2,-3)
dat$beta_format<-paste0(sprintf("%2.2f",dat$beta),
  " (",sprintf("%2.2f",dat$beta_lci)," to ",
  sprintf("%2.2f",dat$beta_uci),")")

# Generate fobj:
fobj2<-genfobj(layout = c("t","t","t","t","t","t","f","t"),
	dat = dat, lwidths = c(0.8,0.4,0.6,0.4,0.6,1,1,0.5))

# Limit the axis
fobj2<-f_axis(fobj2, xlim=c(-1.5,1))

# Add arowheads
fobj2<-f_cutarrows(fobj2)

plotfobj(fobj2)
```


## The header

By default a header with the columns names is used and stored in the *header* element of **fobj**,
which is a list of length 1 with these elements.

```{r}
fobj$header
```

The *header* function can be used to modify the options using all options from [graphics::text()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/text.html)
As an extra element, the *hlayout* can be used to merge columns, i.e. to print a label over more than one column.
And more than one header row can be specified using *headernr*, leading to a list of length > 1.

Let's first just use different names, also including a line separator. 
Note that an empty character has to be included to leave column 1 and 8 empty.
And the y is also modified to place the label it a bit higher.

```{r, fig.height=3.5, fig.width=8}

fobj<-header(fobj = fobj,
  labels = c("","Arm A\nN","Arm A\nmean (sd)","Arm B\nN","Arm B\nmean (sd)",
	"Mean difference\n95% CI","","P-value"),
  y = 0.6)

plotfobj(fobj)
```

We could also merge the label for the effect over the format and forest columns.
  We would then change the layout to include one label twice. And remove the empty label for the forest column.

```{r, fig.height=3.5, fig.width=8}

fobj<-header(fobj = fobj, hlayout = c(1,2,3,4,5,6,6,7),
  labels = c("","Arm A\nN","Arm A\nmean (sd)","Arm B\nN","Arm B\nmean (sd)",
	"Mean difference 95% CI","P-value"))

plotfobj(fobj)
```

In order to also merge to arm labels, we would need to header rows using option *headernr*, leading to a header list with length 2:
As before we can use further [graphics::text()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/text.html) options.

```{r}

fobj<-header(fobj=fobj, hlayout = c(1,2,2,3,3,4,4,5),  headernr = 1,
	labels=c("","Arm A","Arm B","Mean diff (95% CI)","P-value"),
	y=0.9)

fobj<-header(fobj=fobj, hlayout = c(1,2,3,4,5,6,7,8), headernr = 2,
	labels=c("","N","Mean (sd)","N","Mean (sd)","","",""),y=0.3)


fobj$header	
```

```{r, fig.height=3.5, fig.width=8}

plotfobj(fobj)

```

## Boxplots

For a more in-depth presentation of the data in each group, a boxplot can be added.
However, it depends on the input of the actual observations as a data frame *obs* in a long format with 
the numerical outcome value (*value*), the variable name (*variable*, as a factor) and the treatment arm (*arm*, as a factor).

Note that the names of *variable* do not have to be the same as in the summary data (*dat*) but the order has to be kept (i.e. the first level of *variable* must correspond the first row in *dat*).

Boxplot layout can be controlled via helper functions *b_boxplot* and *b_axis* using all options available for [graphics::boxplot()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/boxplot.html) and [graphics::axis()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/axis.html).

```{r, fig.height=3.5, fig.width=8}

fobj<-genfobj(dat = forplotdata, obs = forplotdata_bp,
  layout = c("t","t","t","t","t","b","t","f","t"),
  lwidths = c(0.6,0.4,0.6,0.4,0.6,1,1,1,0.5))

plotfobj(fobj)
```

Adding header gridlines and stripes:

```{r, fig.height=3.5, fig.width=8}

fobj<-gridlines(fobj)

fobj<-stripes(fobj)

fobj<-header(fobj, hlayout = c(1,2,2,3,3,4,5,5,6),  headernr = 1,
	labels=c("","Arm A","Arm B","","Mean difference (95% CI)","P-value"),
	col = c(1,"red","blue",1,1),
	y=0.9)
fobj<-header(fobj, hlayout = c(1,2,3,4,5,6,7,8,9), headernr = 2,
	labels=c("","N","Mean (sd)","N","Mean (sd)","","","",""),
	col=1, y=0.3)

plotfobj(fobj)
```


## Density plots 

Density plots can be specified using layout item *d* and depend on the same data with the observations as the boxplots.

```{r, fig.height=3.5, fig.width=8}

fobj<-genfobj(dat = forplotdata, obs = forplotdata_bp,
  layout = c("t","t","t","t","t","b","d","t","f","t"),
  lwidths = c(0.3,0.4,0.6,0.4,0.6,1,1,1,1,0.5))
 
fobj<-b_axis(fobj, xlim=c(0,9.5))

fobj<-gridlines(fobj)

fobj<-stripes(fobj)

fobj<-header(fobj, hlayout = c(1,2,2,3,3,4,5,6,6,7),  headernr = 1,
	labels=c("","Arm A","Arm B","","","Mean difference (95% CI)","P-value"),
	col = c(1,"red","blue",1,1,1),
	y=0.9)
fobj<-header(fobj, hlayout = c(1,2,3,4,5,6,7,8,9,10), headernr = 2,
	labels=c("","N","Mean (sd)","N","Mean (sd)","","","","",""),
	col=1, y=0.3)

plotfobj(fobj)

```

Options can be change via *d_axis* and *d_lines* using all options available for [graphics::axis()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/axis.html) and [graphics::lines()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/lines.html).

Note that the *lines* element of the *d* items is a nested list with *variable* and *arm*.
Using *d_lines*, line options can be accessed all at once, over all variables for one arm, for all arms and one variable, or for a specific variable-arm combination via *linenr*.
*linenr* is a vector of length two where the first element specifies the variable and the second the arm.

```{r, fig.height=3.5, fig.width=8}
fobj<-genfobj(dat = forplotdata, obs = forplotdata_bp,
  layout = c("t","t","t","t","t","d","t","f","t"),
  lwidths = c(0.3,0.4,0.6,0.4,0.6,1,1,1,0.5))

#all lines:

fobj<-d_lines(fobj=fobj, lw=1.5)
plotfobj(fobj)
```


```{r, fig.height=3.5, fig.width=8}
#only one arm:

fobj<-d_lines(fobj=fobj, linenr=c(NA,2), col=1)
plotfobj(fobj)
```


Different density curves could be added by using *x* and *y* options in *d_lines* (or the *lines* list in the **fobj**).
Note that the y-position has to be shifted by *y.at* for each variable.

## Strip plot for proportions

For binary outcomes and in particular for serious adverse event reporting 
a graphical representation of the proportion in both arms has been recommended.

A strip plot for the proportions can be added via *s[1-9]*, where the number would indicate the number of points in the strip plot and the number of columns in the *dat* (usually two if there are two treatment arms).
The "s" items then contains several *points* elements.

For example:

```{r, fig.height=3.5, fig.width=8}

fobj<-genfobj(dat = forplotdata_prop,
  layout = c("t","t","t","t","t","s2","t","f","t"),
  lwidths = c(0.3,0.4,0.6,0.4,0.6,1.0,1.2,1,0.5))

plotfobj(fobj)
```

Options can be modified via *s_axis*, *s_hline* and *s_points* using all options available for using all options available for [graphics::axis()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/axis.html), [graphics::abline()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/abline.html) and [graphics::points()](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/points.html).

Note that for points, each sub-item has can be specified separately using *pointnr* (e.g. to specify colors).

Left and right borders can be added via *s_borders*.

```{r, fig.height=3.5, fig.width=8}

fobj<-s_axis(fobj=fobj, xlim = c(0,1), 
  at = seq(0,1,by=0.25), labels = seq(0,100,by=25))

fobj<-s_points(fobj=fobj, pch = 16, cex=1.5)

fobj<-s_points(fobj=fobj, pointnr = 1, col = "red")

fobj<-s_points(fobj=fobj, pointnr = 2, col = "blue")

fobj<-s_borders(fobj)

fobj<-gridlines(fobj)

plotfobj(fobj)
```

## Combine plots 

`plotfobj` also works with a list of `fobj` as long as the layout has the same length. If the `lwdiths` vary, the average is taken.


```{r, fig.height=6, fig.width=8}
#define common lwidths
commonlwidths<-c(0.3,0.4,0.6,0.4,0.6,1,1,1,0.5)

#prepare first fobj
fobj1<-genfobj(dat = forplotdata, obs = forplotdata_bp,
  layout = c("t","t","t","t","t","b","t","f","t"),
  lwidths = commonlwidths)

fobj1$setup$lheights[1]<-0.2
fobj1$setup$lheights[3]<-0.2

fobj1<-gridlines(fobj = fobj1)
fobj1<-stripes(fobj = fobj1)

fobj1<-header(fobj = fobj1, hlayout = c(1), headernr = 1,
	labels=c("Continuous variables"),
	col = 1, y = 0.9, font = 2, cex=1.2)

fobj1<-header(fobj = fobj1, hlayout = c(1,2,2,3,3,4,5,5,6),  headernr = 2,
	labels=c("","Arm A","Arm B","","Mean difference (95% CI)","P-value"),
	col = c(1,"red","blue",1,1), y = 0.45, font = 1, cex = 1)

fobj1<-header(fobj = fobj1, hlayout = c(1,2,3,4,5,6,7,8,9), headernr = 3,
	labels=c("","N","Mean (sd)","N","Mean (sd)","","","",""),
	col = 1, y = 0.15, font = 1, cex = 1)


#prepare second fobj
fobj2<-genfobj(dat = forplotdata_prop,
  layout = c("t","t","t","t","t","s2","t","f","t"),
  lwidths = commonlwidths)

fobj2$setup$lheights[1]<-0.2
fobj2$setup$lheights[3]<-0.2

fobj2<-s_axis(fobj = fobj2, xlim = c(0,1), 
  at = seq(0,1,by=0.25), labels = seq(0,100,by=25))

fobj2<-s_borders(fobj = fobj2)

fobj2<-gridlines(fobj = fobj2)
fobj2<-stripes(fobj = fobj2)

fobj2<-header(fobj = fobj2, hlayout = c(1), headernr = 1,
	labels=c("Binary variables"),
	col = 1, y = 0.9, font = 2, cex=1.2)

fobj2<-header(fobj = fobj2, hlayout = c(1,2,2,3,3,4,5,5,6),  headernr = 2,
	labels=c("","Arm A","Arm B","Proportion (%)","Odds ratio (95% CI)","P-value"),
	col = c(1,"red","blue",1,1), y = 0.45, font = 1, cex = 1)

fobj2<-header(fobj = fobj2, hlayout = c(1,2,3,4,5,6,7,8,9), headernr = 3,
	labels=c("","N","n (%)","N","n (%)","","","",""),
	col = 1, y = 0.15, font = 1, cex = 1)


#plot list:
plotfobj(fobj = list(fobj1, fobj2))
```

Combining plots can also be used to get different scaling for the plots of the individual variables.
Assume we would like to present a number of differently scaled continuous variables with boxplots and densities:

```{r, fig.height=5, fig.width=8}
set.seed(1)

fobj<-vector(10,mode="list")

for (i in 1:10) {
  
  #modify the data
	obs<-forplotdata_bp[forplotdata_bp$variable==paste0("var",i),]
	obs$value<-obs$value + runif(1,-10,10)
	obs$variable<-factor(obs$variable)
	
	dat<-cbind(forplotdata[i,1:5],"")
	dat$n2<-paste0(sprintf("%1.1f",mean(obs$value[obs$arm==1]))," (",
    sprintf("%1.1f",sd(obs$value[obs$arm==1])),")")
	dat$n4<-paste0(sprintf("%1.1f",mean(obs$value[obs$arm==2]))," (",
    sprintf("%1.1f",sd(obs$value[obs$arm==1])),")")
	
	#generate individual fplots
	fobji<-genfobj(dat = dat, obs = obs,
		layout = c("t","t","t","t","t","b","t","d"), 
		lwidths = c(0.3,0.4,0.6,0.4,0.6,1,0.1,1),
		lheights = c(0.2,1,0.25))
	
	fobji<-b_axis(fobji, cex.axis=0.8, mgp=c(0,0.3,0), tck=-0.1)
	fobji<-d_axis(fobji, cex.axis=0.8, mgp=c(0,0.3,0), tck=-0.1)
	
	#header for first fplot only
	if (i!=1) { 
		fobji$header<-NULL
	} else {
		
		fobji<-header(fobj = fobji,
      hlayout = c(1,2,2,3,3,4,5,6), headernr = 1,
			labels=c("","Arm A","Arm B","Boxplot","","Density"),
			col = c(1,"red","blue",1,1), y = 0.9)

		fobji<-header(fobj = fobji,
      hlayout = c(1,2,3,4,5,6,7,8), headernr = 2,
			labels=c("","N","mean (sd)","N","mean (sd)","","",""),
			col = 1, y = 0.35)
		
		fobji<-gridlines(fobji,gridnr=2)
	}
	
	#collect
	fobj[[i]]<-fobji
	
}

#adapt top header and bottom footer
fobj[[1]]$setup$lheights[1]<-1
fobj[[length(fobj)]]$setup$lheights[3]<-0.5

#plot the list of fobj
plotfobj(fobj)

```

